# Installation and Setup


If asked to update packages, select "n"
```{r}
BiocManager::install("muscData")

```


We are going to work with a dataset looking at peripheral blood mononuclear cells from patients with lupus treated with IFNB. (https://www.nature.com/articles/nbt.4042)

```{r}
library(muscData)
sce <- Kang18_8vs8()
```

---

## Single Cell Experiment

Single cell experiment objects are the R equivalent of an anndata object and store all of the data for a single cell experiment in a single data structure.

This will return the basic dimensions of the dataset in the form of genes x cells:

```{r}
dim(sce)
```

---

## Three main components of the object

1. **Assays** (count matrices)
2. **colData** (cell-level metadata)
3. **rowData** (gene-level metadata)

```{r}
# Full counts matrix, note that it's sparse to save on size
counts(sce)
```

```{r}
# Cell metadata
head(colData(sce))
```

```{r}
# Gene metadata
head(rowData(sce))
```

---

## Exploring Cell Metadata

```{r}
colData(sce)
```

The `ind` column contains the ID of each individual sample. How many unique samples are there?

Hint: You can access columns of the sce experiment object with the `sce$column-name` notation. Find a function in R that returns unique values. 


```{r}
length(unique(sce$ind))

```

The `stim` column indicates whether cells were stimulated with interferon-β or left as controls. How many cells are in each condition?

Hint: Use the `table()` function.

```{r}
table(sce$stim)
```

Create a cross-tabulation showing the number of cells for each individual in each condition:

Hint: Use the `table()` function and specify the appropriate columns in the sce object using sce$column_name

```{r}

table(Individual = sce$ind, Condition = sce$stim)
```


The `cell` column contains cell type annotations. Which cell type has the most cells?


```{r}
cell_counts <- table(sce$cell)
sort(cell_counts, decreasing = TRUE)
```



## Exploring Gene Metadata

```{r}
head(rowData(sce))
```

The gene metadata contains Ensembl IDs and gene symbols. Find the gene symbol for "ENSG00000087086":

Hint: Remember rowData(sce) is a dataframe and can be subsetted rowData(sce)[rows, columns]. 
You will need an expression to search for the right gene, and return all of the columns

```{r}
rowData
```


```{r}
rowData(sce)[rowData(sce)$ENSEMBL == "ENSG00000087086",]

```


## Subsetting SingleCellExperiment Objects

SCE objects use `sce[genes, cells]` syntax for subsetting.

Make a new sce object with just the cells labeled as B cells

Hint: You may need to filter out NA values in addition to checking if a cell
is a B-cell

```{r}

sce_bcells <- sce[, !is.na(sce$cell) & sce$cell == "B cell"]
sce_bcells
```

Make a new sce object with the B cells from only the stimulated condition

```{r}
stim_bcells <- sce[,sce$stim == "stim"]
stim_bcells
```

Create a subset containing only CD14+ Monocytes from individual 101:

```{r}
cd14_101 <- sce[sce$ind == "101", !is.na(sce$cell) & sce$cell == "CD14+ Monocytes"]

```


## Quality Control Metrics


Find a function that will calculate the sum of all counts for each column

```{r}


```

Find how many genes are expressed in each cell (genes with > 0 counts)
```{r}


```

Add both of the above calculated values to the coldata
```{r}

```

## Preparing for Pseudobulk Analysis

For pseudobulk analysis, we aggregate cells by individual and condition.

Find an R function that will combine the values of the `ind` and `stim` columns
into a new value and save that to a new column in the sce object. 

```{r}




```


For CD14+ Monocytes, how many cells contribute to each pseudobulk sample?

Hint: You can again subset the sce object by sce[rows, ] and use the `table()`
function. Make a subsetted object that only contains the cell type of interest
and then count by the correct column (pseudobulk column) using the `table()`
function.

```{r}


```


Let's convert the single cell experiment object to a seurat object so that we can do the initial workflow on the samples

```{r}


```

---
title: "Pseudobulk.Rmd"
output: html_document
date: "2025-12-05"
---

Copy and paste this Rmarkdown into your previous Rmd that we were working on 
during class Wednesday.


Install the muscat package through BiocManager
```{r}

```


Use the aggregateData function from muscat to generate matrices where you've added
up the counts from each cell population to a single count per sample. Pseudobulking
has been shown to mitigate many of the biases when trying to compare expression 
profiles of individual cells of the same type against individual cells. 

You will want to use the `cell` and `sample_id` column to group and count cells
```{r}

```
  
 
Choose just one of the cell populations at random, and extract out the counts
using assay(sce_object, "name_in_cell_column"). Take a look, for each cell
population we now have a matrix where the rows are genes, and the columns are the
16 samples (8 stimulated, 8 control).
```{r}


```
 
 
 
Make the colData for DESeq2. Remember you must list out the samples in the same
order as the columns in the counts. You should make a new column that has the
metadata about whether that sample is "stim" or "ctl". The rownames should be 
the same as the column names. 

Hint: You can use a combination of colnames(), data.frame(), rownames(), and 
str_split_i() to accomplish this.

```{r}

```
 
 
Load the DESeq2 library and create the DDS object using the counts, Coldata, and
set the design formula as `~ condition`
```{r}


```

Run DESeq2 and create a tibble of the results ordered by pvalue
```{r}

```
 
 
Let's try to recreate their other findings from the single cell data by following
the basic Seurat workflow. Create a new Seurat object using the `CreateSeuratObject()`
function. You will need to provide the counts from the entire single cell experiment
object and the metadata stored in colData.

```{r}


```


Do some basic QC analysis on the data following the Seurat tutorial
```{r}

```


Normalize the data, find highly variable features, scale the data and run PCA.
Again, you may follow the Seurat tutorial and use the suggested starting values. 
```{r}

```


Run FindNeighbors, FindClusters, and RunUMAP. Ensure that you specify a resolution for FindClusters and use
your choice of # of Principal components for FindNeighbors and RunUMAP. 

Create a UMAP plot of the single cell data. What do you observe?


```{r}

```

Use FindAllMarkers to find marker genes for every cluster, and make a dotplot / 
heatmap of the top 5 marker genes for every cluster. 

```{r}


```

Use the single cell experiment object of the same data and try to use singleR to automatically annotate
the cell types. You will need to convert the seurat object back to a single cell object using  `as.SingleCellExperiment`.
There are several tutorials available online for how to run singleR and celldex, it is conceptually quite similar to
celltypist.

```{r}


```


## Session Info

```{r}
sessionInfo()
```