---
title: "analysis"
output:
  pdf_document: default
  html_document: default
---


Load the libraries needed
```{r}
library(DESeq2)
library(tidyverse)
```



Read in the counts matrix using `read.table` and `as.matrix`
```{r, echo=FALSE}
counts <- as.matrix(read.table("all_verse_outputs.tsv", header = TRUE, row.names = "gene_id"))

head(counts)
```

In the last lab, we manually generated a coldata. While convenient for small 
experiments, it would be better for us to directly create the coldata for the 
experiment based on the column names.

Use a combination of `colnames()`, `str_split_i()` and `data.frame()` to make the
coldata. Name the column `condition`.
```{r, echo=FALSE}
coldata <- data.frame(row.names = colnames(counts), condition = str_split_i(colnames(counts), "_", i =1))

coldata

```



Now create a DESeq2 object as you've done in the past to test for DE using the design formula
~ condition

```{r, echo=FALSE}
dds <- DESeqDataSetFromMatrix(countData = counts, colData = coldata, design = ~ condition)
```


Run DESeq2 and extract out the results comparing KO to WT
```{r, echo=FALSE}
dds <- DESeq(dds)
res <- results(dds, contrast = c("condition", "KO", "WT"))
resOrdered <- res[order(res$padj),]
```

Convert the results to a tibble using `as_tibble()` and ensure you keep the rownames.
Use a tidyverse function and report the number of genes that are significant with
a padj < .15
```{r, echo=FALSE}
resOrdered <- as_tibble(resOrdered, rownames = "gene_id")

resOrdered %>% filter(padj < 0.15)

```

Technically, DESeq2 does not ever generate a normalized counts matrix. It performs
normalization and the size factors are incorporated into the model. However, DESeq2
does provide a function to return a matrix of normalized counts. We are often
interested in the normalized counts as visualizing them can help us confirm
the success of our differential expression analysis as well as reveal other
interesting patterns.

Find the appropriate function in DESeq2 and generate a matrix of normalized counts
```{r, echo=FALSE}


```

Take the gene with the lowest padj and plot the normalized counts per sample.
There are many ways to accomplish this. 

I used a combination of `as_tibble()`, `filter`, `pivot_longer()`, `separate_wider_delim()`,
`ggplot`, and `geom_col()`.

```{r, echo=FALSE}


```

This was a gene called significantly differentially expressed, and you can see
that in general, the counts are higher for the WT samples. However, there does
appear to be large variance in the counts for the WT.

This was a real experiment and we noticed that we didn't appear to have many
differentially expressed genes in total but we observed a very striking phenotype.
This led us to believe that we were missing something in our analysis.

If you look at the numbers of the samples, do you see any patterns?

As it turns out, this was an experiment done with mice and the numbers relay
the information on which mice were part of which litter. 840, 841, 842, 842 were
one litter of mice, 857 and 858 were another, and 891 and 892 another. 

Oftentimes, there are other factors that we believe may be affecting gene expression.
We call these covariates and it can be helpful to include them in your model when
testing for differential expression.

Let's redo the analysis and add in the knowledge of our samples.

Update your coldata by adding a new column indicating which litter samples came 
from. Assign the following labels by calling `data.frame` on coldata and adding
a new column called litter. Remember you may make lists in R using `c()`, for example,
c('L1', 'L1') would make a list / vector of two elements.
```{r, echo=FALSE}


```


Let's recreate our DESeq2 object with the new coldata and an updated design
```{r}

```

Re-run DESeq2 and regenerate the results
```{r}

```

Convert your results to a tibble and now report how many genes are significant
at padj < .15.
```{r, echo=FALSE}

```

Re-extract the normalized counts
```{r}


```

Now let's plot the counts for the gene ENSMUSG00000025780.7:
```{r, echo=FALSE}


```

And the same for gene ENSMUSG00000002015.5
```{r, echo=FALSE}



```